---
title: "NatureOfCrime"
author: "Ronnel Salgado"
format: html
editor: visual
---

## Load tidyverse

```{r}
#Load tidyverse
library(tidyverse)
```

## Read data-sets into tibble

```{r}
#read all violent crime csv file into tibble 
allCrimeDataset <- read_csv("natureofcrimeviolenceyemar2023- Table_1a.csv")
#read crime resulting in injury csv file into tibble 
injuryDataet <-  read_csv("natureofcrimeviolenceyemar2023(Table_1b).csv")
#read crime resulting in no injury csv file into tibble 
noInjuryDataet <-  read_csv("natureofcrimeviolenceyemar2023(Table_1e).csv")



#to remove all variables rm(list = ls())


```

## Tidying all violent crime Data

```{r}
#removing and renaming rows/columns
allCrimeTib <- allCrimeDataset %>%
  slice(-c(1:8)) %>%
  select(-c(11:12))  %>%
  slice(-c(1, 2, 3, 4, 7, 8, 10, 11, 13, 14, 15, 16, 17, 18 )) %>%
  rename(Timing = 1,
  Apr2012toMar2013 = 2,
  Apr2013toMar2014 = 3,
  Apr2014toMar2015 = 4,
  Apr2015toMar2016 = 5,
  Apr2016toMar2017 = 6,
  Apr2017toMar2018 = 7,
  Apr2018toMar2019 = 8,
  Apr2019toMar2020 = 9,
  Apr2022toMar2023 = 10
  ) 
  
  #format into long format
  allCrimeTibLong <- allCrimeTib %>%  pivot_longer (
    cols = starts_with("Apr"),
    names_to = "Year",
    values_to = "Percentage"
  )%>%
  mutate(Year = factor(Year), #factorise Year
         Timing = factor(Timing), #factorise Timing
         Percentage = as.numeric(Percentage)) #turn percentages as numbers
```

## 

## Tidying violent crime with injury

```{r}
#removing and renaming rows/columns
injuryTib <- injuryDataet %>%
  slice(-c(1:8)) %>%
  select(-c(11:12))  %>%
  slice(-c(1, 2, 3, 4, 7, 8, 10, 11, 13, 14, 15, 16, 17, 18 )) %>%
  rename(Timing = 1,
  Apr2012toMar2013 = 2,
  Apr2013toMar2014 = 3,
  Apr2014toMar2015 = 4,
  Apr2015toMar2016 = 5,
  Apr2016toMar2017 = 6,
  Apr2017toMar2018 = 7,
  Apr2018toMar2019 = 8,
  Apr2019toMar2020 = 9,
  Apr2022toMar2023 = 10
  ) 
  
  #format into long format
  injuryTibLong <- injuryTib %>%  pivot_longer (
    cols = starts_with("Apr"),
    names_to = "Year",
    values_to = "Percentage"
  )%>%
  mutate(Year = factor(Year), #factorise Year
         Timing = factor(Timing), #factorise Timing
         Percentage = as.numeric(Percentage)) #turn percentages as numbers
```

## Tidying violent crime with no injury

```{r}
#removing and renaming rows/columns
noInjuryTib <- noInjuryDataet %>%
  slice(-c(1:8)) %>%
  select(-c(11:12))  %>%
  slice(-c(1, 2, 3, 4, 7, 8, 10, 11, 13, 14, 15, 16, 17, 18 )) %>%
  rename(Timing = 1,
  Apr2012toMar2013 = 2,
  Apr2013toMar2014 = 3,
  Apr2014toMar2015 = 4,
  Apr2015toMar2016 = 5,
  Apr2016toMar2017 = 6,
  Apr2017toMar2018 = 7,
  Apr2018toMar2019 = 8,
  Apr2019toMar2020 = 9,
  Apr2022toMar2023 = 10
  ) 
  
  #format into long format
  noInjuryTibLong <- noInjuryTib %>%  pivot_longer (
    cols = starts_with("Apr"),
    names_to = "Year",
    values_to = "Percentage"
  )%>%
  mutate(Year = factor(Year), #factorise Year
         Timing = factor(Timing), #factorise Timing
         Percentage = as.numeric(Percentage)) #turn percentages as numbers
```

## Shannon Entropy of all crime types

```{r}
totalPercentage <- sum(allCrimeTibLong$Percentage)
entropyAllCrime <- allCrimeTibLong %>%
  #find probability for each timing
  group_by(Timing) %>%
  mutate(probability = Percentage / totalPercentage) %>%
  ungroup()


entropy <- entropyAllCrime %>%
  #find entropy for each timing
  group_by(Timing) %>%
  summarise(entropy = -sum(probability * log2(probability))) %>%
  ungroup()
  
  #sum(entropyAllCrime$probability * log2(1/entropyAllCrime$probability))

entropy
```

## Linear regression model for all violent crimes

```{r}

#linear regression model
modelAllCrime <- lm(Percentage ~ Timing, data = allCrimeTibLong) 

#display linear model
summary(modelAllCrime) 

#weekend as reference: tib$Timing <- relevel(tib$timing, ref = "During the week")
#morning as reference: tib$Timing <- relevel(tib$timing, ref = "Morning")
```

## Linear regression model of crimes with injury

```{r}
#linear regression model
modelWithInjury <- lm(Percentage ~ Timing, data = injuryTibLong) 

#display linear model
summary(modelWithInjury) 
```

## Linear regression model of crimes with no injury

```{r}
#linear regression model
modelNoInjury <- lm(Percentage ~ Timing, data = noInjuryTibLong) 

#display linear model
summary(modelNoInjury) 
```

## Plot linear regression model for all violent crimes

```{r}
#plot linear model
ggplot(allCrimeTibLong, aes(x = Year, y = Percentage, color = Timing)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Timing), linetype = "solid") +
  labs(
    title = "Linear Regression of all violent crime percentages occured by Year and Timing",
    x = "Year",
    y = "Percentage of Violent Crimes",
    color = "Time of Day"
    
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Plot linear regression model for crimes with injury

```{r}
#plot linear model
ggplot(injuryTibLong, aes(x = Year, y = Percentage, color = Timing)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Timing), linetype = "solid") +
  labs(
    title = "Linear Regression of violent crime percentages occured by Year and Timing",
    x = "Year",
    y = "Percentage of Violent Crimes",
    color = "Time of Day"
    
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Plot linear regression model for crimes with no injury

```{r}
#plot linear model
ggplot(noInjuryTibLong, aes(x = Year, y = Percentage, color = Timing)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Timing), linetype = "solid") +
  labs(
    title = "Linear Regression of violent crime percentages occured by Year and Timing",
    x = "Year",
    y = "Percentage of Violent Crimes",
    color = "Time of Day"
    
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Weekdays vs Weekend Tidy

```{r}
#removing and renaming rows/columns
weekdayWeekend <- allCrimeDataset %>%
  slice(-c(1:8)) %>%
  select(-c(11:12))  %>%
  slice(-c(3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18 )) %>%
  rename(Timing = 1,
  Apr2012toMar2013 = 2,
  Apr2013toMar2014 = 3,
  Apr2014toMar2015 = 4,
  Apr2015toMar2016 = 5,
  Apr2016toMar2017 = 6,
  Apr2017toMar2018 = 7,
  Apr2018toMar2019 = 8,
  Apr2019toMar2020 = 9,
  Apr2022toMar2023 = 10
  ) 
  
  #format into long format
  weekdayWeekendLong <- weekdayWeekend %>%  pivot_longer (
    cols = starts_with("Apr"),
    names_to = "Year",
    values_to = "Percentage"
  )%>%
  mutate(Year = factor(Year), #factorise Year
         Timing = factor(Timing, levels = c("During the week", "At the weekend[note 5]")), #factorise Timing and reorder level
         Percentage = as.numeric(Percentage)) #turn percentages as numbers
```

## Weekday vs Weekend t-test

```{r}
ttestWeekendvsWeekday <- t.test(Percentage ~ Timing, data = weekdayWeekendLong, var.equal = TRUE)

print(ttestWeekendvsWeekday)
```

## Weekday vs Weekend Linear Regression Model

```{r}

#linear regression model
modelWeekdayWeekend <- lm(Percentage ~ Year + Timing, data = weekdayWeekendLong) 

#display linear model
summary(modelWeekdayWeekend) 

#weekend as reference: tib$Timing <- relevel(tib$timing, ref = "During the week")
#morning as reference: tib$Timing <- relevel(tib$timing, ref = "Morning")
```

## Plot weekday vs weekend linear regression model

```{r}
#plot linear model
ggplot(weekdayWeekendLong, aes(x = Year, y = Percentage, color = Timing)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = Timing), linetype = "solid") +
  labs(
    title = "Linear Regression of all violent crime percentages occured by Year and Timing",
    x = "Year",
    y = "Percentage of Violent Crimes",
    color = "Time of Day"
    
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
